---
title: "Comparison of CTCF consensus found in the Gained, Lost and stable binding sites common in the KI and KIKI experiments"
author: "by Kaiqiong Zhao"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: false
    number_sections: true
    toc_depth: 4
    keep_md: true
editor_options: 
  chunk_output_type: console
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 4) 
```


```{r}
library(magrittr)
library(rGADEM)
library(ggplot2)
library(reshape2)
#library(MotIV)
library(BSgenome.Hsapiens.UCSC.hg19)
```


```{r}
# 1- Summary without any extensions

#load("~/Documents/Packages/diffMotif/vignette/Step_2_GADEM_for_individual_clusters/Top_1000_three_clusters/GADEM_up_1000.RData")
load("~/Documents/Packages/diffMotif/vignette/Step_6_1_common_KI_KIKI/Data_1000/GADEM_CommonGain.RData")
gadem_up <- gadem

load("~/Documents/Packages/diffMotif/vignette/Step_6_1_common_KI_KIKI/Data_1000/GADEM_CommonLost.RData")
#load("~/Documents/Packages/diffMotif/vignette/Step_2_GADEM_for_individual_clusters/Top_1000_three_clusters/GADEM_down_1000.RData")
gadem_down <- gadem
#load("~/Documents/Packages/diffMotif/vignette/Step_2_GADEM_for_individual_clusters/Top_1000_three_clusters/GADEM_null_1000.RData")
load("~/Documents/Packages/diffMotif/vignette/Step_6_1_common_KI_KIKI/Data_1000/GADEM_CommonStable.RData")
gadem_null <- gadem

motif11 <- gadem_up@motifList[[1]]
motif21 <- gadem_down@motifList[[1]]
motif31 <- gadem_null@motifList[[1]]


summary_motif <- function(motif1){
  seqID_vec <- rep(NA, length(motif1@alignList))
  seq_vec <- strand_vec <- pval <- seqID_vec
  
  for ( i in 1:length(seqID_vec)){
    seqID_vec[i] <- motif1@alignList[[i]]@seqID
    pval[i] <- motif1@alignList[[i]]@pval
    strand_vec[i] <- motif1@alignList[[i]]@strand
    seq_vec[i] <- motif1@alignList[[i]]@seq
  }
  
  out = list(seq_vec=seq_vec, strand_vec = strand_vec, seqID_vec = seqID_vec)
  return(out)
}

rr1 = summary_motif(motif11)
rr2 = summary_motif(motif21)
rr3 = summary_motif(motif31)
```


Motif identified in the Gained cluster

```{r}
motifs <- getPWM(gadem_up)
seqLogo(motifs[[1]])
```


Motif identified in the Lost cluster
```{r}
motifs <- getPWM(gadem_down)
seqLogo(motifs[[1]])
```


Motif identified in the Stable cluster
```{r}
motifs <- getPWM(gadem_null)
seqLogo(motifs[[1]])
```

```{r, eval= F}

motifs <- getPWM(gadem_up)
seqLogo(motifs[[1]])
seqLogo(consensusMatrix( reverseComplement(DNAStringSet(rr1$seq_vec )), as.prob = TRUE)[1:4,])   #  --- direction-- reverse

seqLogo(getPWM(gadem_down)[[1]])  # ---- direction --- not reverse
seqLogo(consensusMatrix( reverseComplement(DNAStringSet(rr2$seq_vec )), as.prob = TRUE)[1:4,])

seqLogo(getPWM(gadem_null)[[1]]) # ---- direction --- not reverse
seqLogo(consensusMatrix( reverseComplement(DNAStringSet(rr3$seq_vec )), as.prob = TRUE)[1:4,])



seqLogo(motifs[[2]])
see = summary_motif( gadem_up@motifList[[2]])$seq_vec
seqLogo(consensusMatrix( reverseComplement(DNAStringSet(see )), as.prob = TRUE)[1:4,])

seqLogo(motifs[[3]])
#seqLogo(motifs[[4]])
#seqLogo(motifs[[5]])
#seqLogo(motifs[[6]])
#seqLogo(motifs[[7]])




nOccurrences(gadem)[1]

#seqLogo( consensusMatrix(DNAStringSet(seqs_set1), as.prob = TRUE)[1:4,])

```



```{r}
# 2- Load the raw sequences

#out.ranges<- readRDS("/Users/kaiqiongzhao/Documents/Projects_since_2021/WitcherKIKI/Results/out_ranges_July_8.rds")
#out_order <- out.ranges[order(out.ranges$PValue),]

gr_null <- read.delim(file = "~/Documents/Packages/diffMotif/vignette/Step_6_1_common_KI_KIKI/Data_1000/CommonStable_1000.txt" ) # 903
# 3- Extension based on Biostrings::getSeq


gr_down = read.delim(file = "~/Documents/Packages/diffMotif/vignette/Step_6_1_common_KI_KIKI/Data_1000/CommonLost_1000.txt" ) # 986
gr_up = read.delim(file = "~/Documents/Packages/diffMotif/vignette/Step_6_1_common_KI_KIKI/Data_1000/CommonGain_1000.txt" ) # 894

Seqs_up=Biostrings::getSeq(Hsapiens, GRanges(gr_up) + 60 )
Seqs_down=Biostrings::getSeq(Hsapiens, GRanges(gr_down) + 60 )
Seqs_null=Biostrings::getSeq(Hsapiens, GRanges(gr_null) +60 )  
```

```{r}

# 4- The main step: Use the extended sequences, mapped on the rr1$seq_vec, i.e. the core motif identified from the original rGADEM

seq_set_extract <- function(rr1, motif11, Seqs_up,bp.use = 21, reverse = FALSE, rrWidth=13, start = 2, end = 14){
  ext = (bp.use-1)/2 - (rrWidth-1)/2
  #if(width(rr1$seq_vec) %>% unique()==13){  # up sequences identified by rGADEM
  
  #}
  #if(width(rr1$seq_vec) %>% unique() ==18){ # down sequences
  #  start = 1; end = 11; center = 6
  #}
  #if(width(rr1$seq_vec) %>% unique() ==14){ # null sequences
  #  start = 3; end = 13; center = 8
  #}
  seqs_set <- rep(NA, length(rr1$strand_vec))
  
  for ( i in 1:length(rr1$strand_vec)){
    if(rr1$strand_vec[i] == "+"){
      seq_nn <- Seqs_up[motif11@alignList[[i]]@seqID] #the raw input long sequence
      #rr1$seq_vec[i]
      oo = gregexpr(DNAString( rr1$seq_vec[i]), seq_nn)
      
      seqs_set[i]<- subseq(seq_nn , start= oo[[1]][1] + (start-1) - ext, end = oo[[1]][1]+(end-1) + ext )
      
      # id_now = rr1$seqID_vec[i]
      #seqs_set[i]<- subseq(Seqs_up[id_now],
      ##   start= motif11@alignList[[i]]@pos+(start-1) - ext ,
      #  motif11@alignList[[i]]@pos+(end-1) + ext  )
    }
    if(rr1$strand_vec[i] == "-"){
      seq_nn <- Seqs_up[motif11@alignList[[i]]@seqID] #the raw input long sequence -- this is in the reverse strand
      
      #oo = gregexpr(DNAString( rr1$seq_vec[i])%>% reverseComplement(), seq_nn)
      #if(oo[[1]][1] != -1){
      
      #seqs_set[i]<-subseq(seq_nn %>% reverseComplement(), start= width(seq_nn)-oo[[1]][1]+1-10 -ext-(width(rr1$seq_vec[i])-end) , #width(seq_nn)-oo[[1]][1]+1 + ext -(width(rr1$seq_vec[i])-end) )
      
      
      # Modify the reverse strand extraction-- July 1 reverse the original input strand instead --> advantage: less tuning for start and end for the - strand
      
      oo = gregexpr(DNAString(rr1$seq_vec[i]), seq_nn %>% reverseComplement())
      if(oo[[1]][1] != -1){
        
        seqs_set[i]<- subseq(seq_nn %>% reverseComplement() , start= oo[[1]][1] + (start-1) - ext, end = oo[[1]][1]+(end-1) + ext )
        
        
        
        # if(width(rr1$seq_vec) %>% unique()==18){
        
        #  subseq(seq_nn , start= oo[[1]][1] + (start-1) - ext, end = oo[[1]][1]+(end-1) + ext ) %>% reverseComplement()
        #seqs_set[i]<-subseq(seq_nn %>% reverseComplement(), start= width(seq_nn)-oo[[1]][1], width(seq_nn)-oo[[1]][1]+1+11 )
        #}
        
      }
    }
    
  }
  seqs_set <- seqs_set[!is.na(seqs_set)]
  
  if(reverse){
    res =consensusMatrix( reverseComplement(DNAStringSet(seqs_set )), as.prob = TRUE)[1:4,]
    
    res_counts = consensusMatrix( reverseComplement(DNAStringSet(seqs_set )), as.prob = FALSE)[1:4,]
  }else{
    res =consensusMatrix(DNAStringSet(seqs_set ), as.prob = TRUE)[1:4,]
    res_counts =consensusMatrix(DNAStringSet(seqs_set ), as.prob = FALSE)[1:4,]
  }
  res_melt1<-melt(res)
  colnames(res_melt1)<- c("Nucleotide", "Position", "Freq.")
  return(list(res = res,res_counts = res_counts, res_melt = res_melt1, seqs_set=seqs_set)) 
  # April 5, 2021, extracting the raw sequence matrix as well
}

res_ext1 <- seq_set_extract(rr1, motif11, Seqs_up, bp.use = 61, reverse = TRUE, rrWidth = 11, start = 2, end = 12)
#-------
# for the up cluster -- the orignal rGADEM identified sequences of length 1-14, center is on position 8
#----------------

# start: start in the original subsequences rr1$seq_vec[i]
# end: end in the original subsequences rr1$seq_vec[i]
# make everything symetric; facilitate any follow-up manipulation for the reverse strand

# For the reverse = TRUE sequences, carefully select the start and end

res_ext2 <- seq_set_extract(rr2, motif21, Seqs_down, bp.use = 61, reverse = FALSE, rrWidth = 11, start = 1, end = 11)
# Down cluster -- 16 bp, center is on position 8
res_ext3<- seq_set_extract(rr3, motif31, Seqs_null, bp.use = 61, reverse = FALSE, rrWidth = 11, start = 3, end = 13)
```



# Frequency Difference Plot


```{r}
res1 = res_ext1$res_melt
res2 = res_ext2$res_melt
res3 = res_ext3$res_melt

```




```{r}

res1_21<- res1
res2_21<- res2
res3_21<- res3



res_tab = cbind(res1_21, res3_21[,3], res2_21[,3])
colnames(res_tab)[3:5] <- c("Gained", "Stable", "Lost")
rownames(res_tab) = NULL
#pander::pander(res_tab)

```



## **Lost** v.s. **Stable**

```{r}

res_tab$`lost-stable` <- res_tab$Lost - res_tab$Stable
res_tab$`Gained-stable` <- res_tab$Gained - res_tab$Stable
#setwd("~/Documents/Packages/diffMotif/vignette/Step_6_1_common_KI_KIKI/Results")
#write.csv(res_tab, "Common_1000_freqTab_61bp.csv")

ggplot(res_tab, aes(Position, `lost-stable`))+
  geom_point(stat = "identity" , aes(colour = Nucleotide, shape = Nucleotide))+
  geom_vline(xintercept = 31, 
             color = "blue", size=0.5, linetype = 3)+
    geom_hline(yintercept = 0.1, linetype = 2, size=0.5, color = "gray") +
     geom_hline(yintercept = -0.1, linetype = 2, size=0.5, color = "gray") +
  geom_vline(xintercept = 25, linetype = 2, size=0.5) +
    geom_vline(xintercept = 48, linetype = 2, size=0.5) +
  ggtitle("Lost cluster v.s Stable Cluster")+
  ylab("Freq. differences")
```

Fig 18: Difference of nucleotide differences between the **Lost** and **Stable** clusters. Black dashed lines (at position 25 and 48) are the boundaries of the region with frequency differences greater than 10% between the Lost cluster and the stable cluster

Table: Nucleotide frequences at positions with Freq. differences greater than 10% between the **Lost** and **Stable** clusters.

```{r}
see = order(abs(res_tab$`lost-stable`), decreasing = T)
res_tab_ord = res_tab[see,]
ss = res_tab_ord
rownames(ss) =NULL
ss = ss[abs(ss$`lost-stable`)>0.1,]
print(ss)
#pander::pander(ss , split.table=Inf)
```




## **Gained** v.s. **Stable**

```{r}
ggplot(res_tab, aes(Position, `Gained-stable`))+
  geom_point(stat = "identity" , aes(colour = Nucleotide, shape = Nucleotide))+
 geom_vline(xintercept = 31, 
             color = "blue", size=0.5, linetype = 3)+
    geom_hline(yintercept = 0.1, linetype = 2, size=0.5, color = "gray") +
     geom_hline(yintercept = -0.1, linetype = 2, size=0.5, color = "gray") +
  geom_vline(xintercept = 33, linetype = 2, size=0.5) +
    geom_vline(xintercept = 47, linetype = 2, size=0.5) +
  ggtitle("Gained cluster v.s Stable Cluster")+
  ylab("Freq. differences")+
  coord_cartesian(ylim = c(-0.2, 0.45))
```

Fig 19: Difference of nucleotide differences between the **Lost** and **Stable** clusters. Black dashed lines (at position 33 and 47) are the boundaries of the region with frequency differences greater than 10% between the Gained cluster and the Stable cluster.

Table: Nucleotide frequences at positions with Freq. differences greater than 10% between the Gained cluster and the Stable cluster.


```{r}
see = order(abs(res_tab$`Gained-stable`), decreasing = T)
res_tab_ord = res_tab[see,]
ss = res_tab_ord
rownames(ss) =NULL
ss = ss[abs(ss$`Gained-stable`)>0.1,]
print(ss)
#pander::pander(ss , split.table=Inf)

```


# Plots of the p-values from the Chi-square test


## Lost vs Stable

```{r Chi-square test}

#  Lost vs. Stable
pval_vec <- rep(NA, 61)

for(posnow in 1:61){
  temp_tab <- data.frame(lost = res_ext2$res_counts[,posnow], stable= res_ext3$res_counts[,posnow])

  tryCatch({pval_vec[posnow] <- chisq.test(temp_tab)$p.value}, 
           warning=function(w) print(posnow))
  
}

pval_vec_plot <- pval_vec
pval_vec_plot[which(pval_vec_plot==0)] <- 10^(-300)

#setwd("~/Documents/Packages/diffMotif/vignette/Step_6_1_common_KI_KIKI/Results")
#pdf("Chi_square_p_vals_Common_1000_Lost_Stable.pdf", height = 5, width = 8)
plot(-log10(pval_vec_plot), pch = 19, xlab = "Position",
     ylab = "-log10(p-value)", 
     main = "Lost cluster v.s Stable Cluster",
     ylim = c(0, 100))
abline(v = 31, col = "blue", lty = 4)
abline(v = 25, lty = 2, col = 1)
abline(v = 48, lty = 2, col = 1)
#dev.off()
```

## Gained vs. Stable

```{r}

#  Gained vs. Stable

pval_vec1 <- rep(NA, 61)

for(posnow in 1:61){
  temp_tab <- data.frame(gained = res_ext1$res_counts[,posnow], stable= res_ext3$res_counts[,posnow])
  tryCatch({pval_vec1[posnow] <- chisq.test(temp_tab)$p.value}, 
           warning=function(w) print(posnow))
}


pval_vec_plot <- pval_vec1
pval_vec_plot[which(pval_vec_plot==0)] <- 10^(-300)

#pval_vec1[which(pval_vec1==0)] <- 10^(-300)

#pdf("Chi_square_p_vals_KIKI_Gained_Stable.pdf", height = 5, width = 8)
plot(-log10(pval_vec_plot), pch = 19, xlab = "Position",
     ylab = "-log10(p-value)", main = "Gained cluster v.s Stable Cluster",
     ylim = c(0, 100))
abline(v = 31, col = "blue", lty = 4)
abline(v = 25, lty = 2, col = 1)
abline(v = 48, lty = 2, col = 1)
#dev.off()
```

```{r}


p_vals_KI <- data.frame("Lost.stable" = pval_vec, "Gained.stable" = pval_vec1)
setwd("~/Documents/Packages/diffMotif/vignette/Step_6_1_common_KI_KIKI/Results")
write.csv(p_vals_KI, file="Chi_square_p_vals_Common_1000.csv")


#plot(-log10(pval_vec_plot), pch = 19)

#plot(-log10(p.adjust(pval_vec_plot)), pch = 19)

```




# Consensus in the identified window

```{r, eval = T}
# check
seqs_set1 <-DNAStringSet(res_ext1$seqs_set) %>% reverseComplement()  # A sequence of length 61
seqs_set2 <- res_ext2$seqs_set
seqs_set3 <- res_ext3$seqs_set



# Matching this to the original input
matchID1 <- rr1$seqID_vec
matchID2 <- rr2$seqID_vec
matchID3 <- rr3$seqID_vec

seqs_set1_sub <- seqs_set1
seqs_set2_sub <- seqs_set2
seqs_set3_sub <- seqs_set3



for( i in 1:length(seqs_set1)){
  seqs_set1_sub[i] <- subseq(seqs_set1[i], 25, 48)
}
for( i in which(!is.na(seqs_set2))){
  seqs_set2_sub[i] <- subseq(seqs_set2[i], 25, 48)
}
for( i in 1:length(seqs_set3)){
  seqs_set3_sub[i] <- subseq(seqs_set3[i], 25, 48)
}
```


## Gained 


```{r, fig.width = 18, fig.height =5}
# Check

seqLogo( consensusMatrix(DNAStringSet(seqs_set1_sub) , as.prob = TRUE)[1:4,])


#df1 <- as.data.frame(gr_up)
#df2 <- as.data.frame(gr_down)
#df3 <- as.data.frame(gr_null)

```


## Stable


```{r, fig.width = 18, fig.height =5}
# Check


seqLogo( consensusMatrix(DNAStringSet(seqs_set3_sub), as.prob = TRUE)[1:4,])


#df1 <- as.data.frame(gr_up)
#df2 <- as.data.frame(gr_down)
#df3 <- as.data.frame(gr_null)

```


## Lost


```{r, fig.width = 18, fig.height =5}
# Check


seqLogo( consensusMatrix(DNAStringSet(seqs_set2_sub[!is.na(seqs_set2_sub)]), as.prob = TRUE)[1:4,])


#df1 <- as.data.frame(gr_up)
#df2 <- as.data.frame(gr_down)
#df3 <- as.data.frame(gr_null)

```



```{r, eval = F}
# check
seqs_set1 <-DNAStringSet(res_ext1$seqs_set) %>% reverseComplement()  # A sequence of length 61
seqs_set2 <- res_ext2$seqs_set
seqs_set3 <- res_ext3$seqs_set


# Check

seqLogo( consensusMatrix(DNAStringSet(seqs_set1), as.prob = TRUE)[1:4,])
seqLogo( consensusMatrix(DNAStringSet(seqs_set2[!is.na(seqs_set2)]), as.prob = TRUE)[1:4,])
seqLogo( consensusMatrix(DNAStringSet(seqs_set3), as.prob = TRUE)[1:4,])

# Matching this to the original input
matchID1 <- rr1$seqID_vec
matchID2 <- rr2$seqID_vec
matchID3 <- rr3$seqID_vec

seqs_set1_sub <- seqs_set1
seqs_set2_sub <- seqs_set2
seqs_set3_sub <- seqs_set3



for( i in 1:length(seqs_set1)){
  seqs_set1_sub[i] <- subseq(seqs_set1[i], 25, 48)
}
for( i in which(!is.na(seqs_set2))){
  seqs_set2_sub[i] <- subseq(seqs_set2[i], 25, 48)
}
for( i in 1:length(seqs_set3)){
  seqs_set3_sub[i] <- subseq(seqs_set3[i], 25, 48)
}

# Check

seqLogo( consensusMatrix(DNAStringSet(seqs_set1_sub) , as.prob = TRUE)[1:4,])
seqLogo( consensusMatrix(DNAStringSet(seqs_set2_sub[!is.na(seqs_set2_sub)]), as.prob = TRUE)[1:4,])
seqLogo( consensusMatrix(DNAStringSet(seqs_set3_sub), as.prob = TRUE)[1:4,])


df1 <- as.data.frame(gr_up)
df2 <- as.data.frame(gr_down)
df3 <- as.data.frame(gr_null)

```




```{r, eval=TRUE}
knitr::knit_exit()
```
